---
title: "Statistical Inference Course Project - Part 1"
author: "Zohar Peleg"
date: "Tuesday, August 12, 2014"
output: 
  word_document
---

This report covers the simulation excercize from the first part of the statistical inference course project. The report does not show the full R code. The complete reproduceable code can be found at my github account at https://github.com/zoharpeleg/Statistical-Inference-Course-Project

In this excercise I have investigated the properties of the distribution of the mean of 40 random exponential variables, using R simulation of the exponential distribution. 
The simulation is using the *rexp(n,lambda)*, where *lambda* is the rate parameter, and *n* is the sample size.
I used *lambda=0.2* and *n=40* for all simulations,as required by the project instructions.
In order to get an idea about the distribution of the sample's properties, I have repeated the simulation for a large number of times *(nosim=1000)*.

Setting the simulation parameters:
```{r,eval=TRUE}
lambda<-0.2 ; n<-40 ; nosim<-1000 ; set.seed(9876543)
```

The simulated samples are stored in a matrix, where each row contains one sample of 40 random exponential variables.
The mean and standard-deviation were calculated for each sample, and stored in two vectors - *meanx*, and *sdx* respectively:
```{r,eval=TRUE}
x<-matrix(rexp(n*nosim,lambda),nosim) #nosim samples of n observations
meanx<-apply(x, 1, mean)              #the means of all samples        
sdx<-apply(x, 1, sd)                  #the std-deviations of all samples
```
1. Showing where the distribution is centered at, and comparing it to the theoretical center of the distribution:

The average mean of all the samples is:
```{r,eval=TRUE}
mean(meanx) 
```
compared with the theoretical mean of the exponential distribution, which is $\mu=1/\lambda=`r 1/lambda`$

2. Showing how variable the sample mean is, and comparing it to the theoretical variance of the distribution:

The calculated variance of the sample means is:
```{r,eval=TRUE}
var(meanx)
```

compared with the theoretical variance of the sample mean, for sample-size of *n*, which is  
$s^2=\sigma^2/n=(1/\lambda)^2/n=5^2/40=`r 5^2/40`$

3. Showing that the distribution is approximately normal:

a. By plotting a normalized histogram of the sample means, compared to the density function curve of normal distribution of $\mu=1/\lambda$ and $\sigma=1/\lambda/\sqrt{n}$:

```{r}
hist(meanx, breaks=40, col="green", freq=F, main="Histogram of meanx vs normal curve")  #plot histogram
xaxis<-seq(min(meanx),max(meanx),length=100) #prepare the x-axis 
lines(xaxis,
      dnorm(xaxis,1/lambda,sd=1/lambda/sqrt(n)),
      type="l", col="red", lwd=2)  #add ref normal curve
```

b. By plotting a Q-Q plot of the sample means compared to random normal sample with the same expected parameters as above:

```{r}
qqplot(rnorm(1000,1/lambda,sd=1/lambda/sqrt(n)),meanx, pch=20, cex = .5, col="black", main="Q-Q plot: meanx vs. normal dist.", xlab="normal")
abline(0,1, col = "red",lwd=1,lty=2)
```

4. Evaluation of the coverage of the confidence interval for $1/\lambda$ using $\overline{X}\pm 1.96*s/\sqrt{n}$:

With the sample's mean and standard deviation, I have calculated the interval's lower limit and upper limit for each sample:
```{r,eval=TRUE}
ul<-meanx+sdx*1.96/sqrt(n) #upper limit vector
ll<-meanx-sdx*1.96/sqrt(n) #lower limit vector
```

Now it is possible to evaluate the percentage of good intervals, which contain the *mean=`r 1/lambda`*, as expected:
```{r,eval=TRUE}
paste(mean(1/lambda<ul&1/lambda>ll)*100,"% of the intervals contained the population mean",sep="")
``` 

Repeating this for many times, I have found that the coverage was consistently around 92%. 
I tried other sample sizes and figured out that the coverage increases as the sample size increases. So I ran the previous code in a loop, sequencing through multiple sample sizes to investigate the effect of the sample size on the coverage. The results are shown in the following figure:
```{r}
nvals <- seq(20, 500, by = 5)
coverage <- sapply(nvals, function(nx) {
        x<-matrix(rexp(nx*1000,0.2),1000)            
        meanx<-apply(x, 1, mean) 
        sdx<-apply(x, 1, sd)        
        ul<-meanx+sdx*1.96/sqrt(nx)
        ll<-meanx-sdx*1.96/sqrt(nx)
        mean(ll < 1/lambda & ul > 1/lambda)
})
plot(nvals,coverage, pch=20, cex=0.5, type="l", lwd=2, xlab="sample size",main="coverage of exponential(0.2) \n as a function of sample size")
lines(smooth.spline(nvals, coverage, spar=0.6), col="red", lwd=2)
```

From the plot above I cnclude that the sample size of 40 observation does not contain enough information as needed to determine the confidence interval according to CLT. 
We see asymptotic convergence towards 95% as the sample size increases. It takes about 250-450 observations to get close to 95%.

I also tried replacing the normal CLT interval with t-interval, by replacing the 1.96 with qt(0.975,n-1) = `r qt(0.975,n-1)`. The coverage increased to about 93%, which is about 0.5%-1% improvement.
